---
title: "exam1"
author: "Mathias Kold"
date: "2024-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exam 1 - OLS and heteroskedasticity

Look at the following model for bank employees wage:
$$log(salary)=\beta_0+\beta_1educ+\beta_2log(salbegin)+\beta_3male+\beta_4 minority+u$$
where salary is yearly wage (in 1000 US dollars), educ is education measured in number of years, salbegin is the starting salary (in 1000 US dollars) for the person's first position in the same bank, male is a dummy variable for gender, minority is one dummy variable indicating whether one belongs to a minority.

## 1 - Estimate the model using OLS. Comment on the output and interpret the results
```{r}
options(scipen = 999)
library(readr)
data1 <- read_csv("data1.csv")
model <- lm(log(salary) ~ educ + log(salbegin) + male + minority, data = data1)
summary(model)
```
From the estimated model it can be seen that the education variable has a value of 0.02327 which means that one extra year of education will raise the salary by approximately 2.3%. The male variable has a value 0.04816 which means that being a man will raise your salary by approximately 4.8%. The minority variable has a value of -0.04237 which means that minorities approximately will have a 4.2% lower salary than people who are not minorities. The salbegin variable have a different interpretation because it is in log form. It has a value of 0.82180 which means that an increase of 1% in your starting salary will make your salary approximately 0.82% higher.

The intercept is 0.84868 which is the value of the dependent variable, in this case salary, when all other variables have a value of 0. So in this case 0.84868 is the expected value of the salary for a person with zero education, without a starting salary and someone who is not a male or a minority.

It can also be seen that all the variables are statistically significant at a 5% significance level due to p-values < 0.05 but only education and log(salbegin) are statistically significant at a 1% significance level. 

## 2 - Perform graphical model checking.
```{r}
plot(model)
```


## 3 - Test for heteroskedasticity using the Breusch-Pagan test and the special edition of the White test.
```{r}
r=residuals(model)
res = r^2
summary(lm(res~educ + log(salbegin) + male + minority, data = data1))
```
Because the p-value is under 0.05 we reject the null hypothesis meaning that there is heteroskedasticity in the model.

We can now calculate the LM test, where we have 474 observations and $R^2=0.1231$:
```{r}
LM = 0.1231*474
LM
```
We can then calculate the p-value of chi-square $\chi^2_k$:
```{r}
1-pchisq(LM,4)
```


```{r}
library(lmtest)
bptest(model)

```


## 4 - Calculate robust standard errors for the model and compare with the results in question 1.

## 5 - Test the hypothesis $H_0:\beta_2 = 1$ against the alternative $H_1: \beta_2 \neq 1$.

## 6 - Test the hypothesis $H_0: \beta_3 = \beta_4 = 0$

## 7 - Estimate the model using FGLS and comment on the results.

## 8 - Has the FGLS estimation taken into account all the heteroskedasticity?
